Evolution of \\ ice sheet geometry \\ using Stokes dynamics

### 2021-06-11 talk/talk.tex
Edits on slides.

\vspace{-1mm}
\begin{itemize}
\item glacier geometry which evolves according to climate
\item take simplest such problem (grounded, nonsliding, isothermal)
\item for simplicity, assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item climatic mass balance $a(x,y)$ \hfill $\gets$ \emph{$a>0$ for snowing}
    \item bed elevation $b(x,y)$
    \end{itemize}
\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow
\item $\therefore$ must solve an equation for the surface elevation
\item the surface kinematical equation (SKE) holds \emph{on} the ice
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s &= a & &\text{\emph{evolving}} \\
\bu|_s \cdot \bn_s + a &= 0 & &\text{\emph{steady state}}
\end{align*}

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
    \item \emph{not} on the ice: $a\le 0$ for $t>0$ (or in steady state)
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item the steady IGP free-boundary problem for $s(x,y)$ on $\Omega$
\item it is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a Stokes problem for $\bu,p$:}

@@

    \begin{itemize}
    \item plus Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \item \emph{count the 3 nonlinearities}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item the evolving IGP free-boundary problem for $s(x,y,t)$ on $\Omega \times [0,T]$
\item consider one step of backward Euler scheme
    \begin{itemize}
    \item solve for $s,\bu,p$ at new time $t^\ell$
    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$, and write $s$ for $s^\ell$
    \end{itemize}
\item again: NCP for $s$ coupled to a Stokes problem for $\bu,p$:

@@

    \begin{itemize}
    \item \alert{pretty similar \dots mostly suffices to consider steady IGP}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Brinkerhoff et al (2017) (\emph{upwinding limits time-step?})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
    \item neither scalable (\emph{single level})
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

@@

\medskip
    \item my goal is to do better
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
\end{itemize}

@@

\begin{center}
\mbox{\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.2\textwidth]{figs/elephant.png}}
\end{center}

>>>>

\vspace{-1mm}
\begin{itemize}
\item glacier geometry which evolves according to climate
\item take simplest such problem (grounded, nonsliding, isothermal)
\item for simplicity, assume time-independent data on fixed $\Omega \subset \RR^2$:
    \begin{itemize}
    \item (net) climatic mass balance $a$ \hfill $\gets$ \emph{$a>0$ for accumulation}
    \item bed elevation $b$
    \end{itemize}
\item \emph{to compute}: surface elevation $s$ on $\Omega$ and icy domain $\Lambda_s \subset \RR^3$
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item glaciers flow
\item $\therefore$ must solve an equation for the surface elevation
\item the surface kinematical equation (SKE) holds \emph{on} the ice
\begin{align*}
\frac{\partial s}{\partial t} - \bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{evolving}} \\
-\bu|_s \cdot \bn_s - a &= 0 & &\text{\emph{steady state}}
\end{align*}

    \begin{itemize}
    \item $\bn_s = \left<-s_x,-s_y,1\right>$  \, is an upward normal to the ice surface
    \item where \emph{not} on the ice: $a\le 0$
    \end{itemize}
\end{itemize}

@@

\begin{itemize}
\item the steady IGP free-boundary problem, for $s(x,y)$ on $\Omega$, is a \alert{nonlinear complementarity problem (NCP) for $s$ which is coupled to a non-Newtonian Stokes problem for $\bu,p$:}

@@

    \begin{itemize}
    \item Glen-law effective viscosity with $\text{p}=(1/\text{n})+1(=4/3)$:
      $$\nu_\eps = \frac{1}{2} B_n \left(|D\bu|^2 + \eps\, D_0^2\right)^{(\pp-2)/2}$$
    \item \emph{count the 3 nonlinearities}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item solve one step of backward Euler for $s,\bu,p$ at new time $t^\ell$
    \begin{itemize}
    \item $\frac{\partial s}{\partial t} \approx \frac{s^\ell-s^{\ell-1}}{\Delta t}$ in evolving SKE, and write $s$ for $s^\ell$
    \end{itemize}
\item again get an NCP coupled to a Stokes problem:

@@

    \begin{itemize}
    \item \alert{similar \dots mostly suffices to consider steady IGP}
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item almost no one is solving the implicit IGP
    \begin{itemize}
    \item except Brinkerhoff et al (2017) (\emph{upwinding, semi-coupled})
    \item and Wirbel \& Jarosch (2020) (\emph{semi-coupled, artificial ice layer})
    \item neither scalable (\emph{single level}) nor unconditionally stable
    \end{itemize}
\item what are people doing instead?
    \begin{itemize}
    \item \alert{explicit time-stepping with $s \ge b$ enforced by truncation}
    \item often with shallow approximations, e.g.~PISM run below

@@

\medskip
    \item \emph{my goal}: unconditionally-stable implicit with Stokes dynamics
    \end{itemize}
\end{itemize}
\end{frame}

@@

\begin{itemize}
\item to compute $\Phi(s)$: build $\Lambda_s$, solve weak-form Stokes problem
    $$F_{\Lambda_s}(\bu,p)[\bv,q] = \int_{\Lambda_s} 2 \nu_\eps D\bu : D\bv - p \Div\bv - (\Div\bu) q - \rhoi \bg \cdot \bv\,d\bx = 0,$$
extract trace $\bus$, then extend $\Phi(s) = -\bus \cdot \bn_s$ by zero to $\Omega$
\item regarding Stokes problem for fixed geometry:
    \begin{itemize}
    \item well-posed over $W_0^{1,\pp}(\Lambda_s)^3 \times L^\qq(\Lambda_s)$ (Jouvet \& Rappaz 2011)
    \item optimal solver exists (Isaac et al 2015)
    \end{itemize}
\item $\therefore$ \, $\phi(s)$ is well-defined if $s$ is piecewise $C^1$
    \begin{itemize}
    \item a regularity conjecture
    \end{itemize}
\end{itemize}

@@

\begin{center}
\includegraphics[width=0.7\textwidth]{figs/idoaction.png} \hfill \includegraphics[width=0.15\textwidth]{figs/elephant.png}
\end{center}
